Morse Theory

### 2020-11-29 paper.tex
Update paper.tex
about non-orientable

>>>>

\textbf{Non-orientable 2-manifolds without boundary}\\

We will consider in this section Reeb graphs of Morse functions over non-orientable 2-manifolds with boundary. In particular, number of nodes, contours and loops in such graphs. \\
All connected non-orientable 2-manifolds without boundary could be expressed as connected sum of  $g$ copies of projective planes $N=P\neq P..\neq P$, where $g$ coincides with the genus of the surface.\\
For the further proofs we have to recall that Euler characteristic of orientable surfaces can be calculated as $\chi=2-2g$ and for non-orientable it is $\chi=2-g$. Also we have to notice that we can make a 2-sheeted cover of a non-orientable 2-manifold, where each sheet is orientable. In future we will call this $doubling$. \\
Let's consider the process of doubling. We obtain 2-sheeted cover of $N$ (non-orientable 2-manifold) by doubling each point $x\in N$ to two points $x'$ and $x''$. We can consider them as a points lying in the neighborhood of $x$ on two opposite sides of $N$. Union of such points $x'$ and $x''$ will be space that is connected and orientable 2-manifold $M$. Due to the doubling Euler characteristic of the surface will be doubled, $\chi_M=2\chi_N$. \\
Now we have to calculate genus of $M$. Since it is orientable:\\ $g_M=\frac{2-\chi_M}{2}=1-\frac{\chi_M}{2}=1-\chi_N=1-(2-g_N-1)=g_N-1$. \\
Therefore, by doubling we decrease genus by 1. \\
In the table on the fig. \ref{fig:Table} one can find Euler characteristicand genus number of basic non-orianatble manifolds. Note that $K$ is a Klein bottle, $T$ represents torrus,$S$ sphere and $P$ is a projective plane. 
\begin{figure}[h!]
\center{\includegraphics[scale=0.9]{char_table.png}}
\caption{Euler characteristic and genus of some non-orientable manifolds and their doublings}
\label{fig:Table}
\end{figure}
On the table fig. \ref{fig:Table} one can observe that relation of Euler characteristic and genus number of non-orientable manifolds and their doublings coincides with the dependence introduced above. \\
Now we can move on to describing Reeb graphs of non-orientable 2-manifolds without boundary. As an example let's consider Klein bottle, its Reeb graph, and Reeb graph of its doubling surface on fig.\ref{fig:Klein}. \\
\begin{figure}[h!]
\center{\includegraphics[scale=0.45]{Klein_bottle.png}}
\caption{Reeb graph of Klein bottle and its doubling}
\label{fig:Klein}
\end{figure}
As we see on the fig. \ref{fig:Klein} Reeb graph of the surface drastically changes after doubling, especially number of nodes and contours.  Let's learn how do nodes of each degree change under this doubling. \\
1-degree nodes are maxima or minima, after doubling number of these nodes is doubled and also number of contours is doubled as well. \\
2-degree nodes exist in Reeb graphs of non-orientable surfaces, despite the fact that in orientable surfaces they could be just collapsed. It is transformed into 4-degree node after doubling. The reason is that the level set containing a critical point of index 1 is a figure-8 which may contain an orientation-reversing cycle. \\
3-degree nodes are saddle points that just transfer into two such nodes in its doubling. \\
One could observe the above on the fig. \ref{fig:Nodes}.  
\begin{figure}[h!]
\center{\includegraphics[scale=0.3]{Nodes.png}}
\caption{Correspondence between nodes in $N$ and its doubling $M$}
\label{fig:Nodes}
\end{figure}
Let $f: N \rightarrow R$ is a Morse function that induces function $f_0: M \rightarrow R$. $f_0$ is not a Morse function, since all critical points of $f$ correspond to two critical points of $f_0$, so they are not distinct. However, critical points of $f_0$ are still non-degenerate, that gives as an opportunity to consider $R(f_0)$ (Reeb graph of $f_0$) as previously. \\
Let $e$ be number of arcs in $R(f)$, $n_{1,3}$ is number of 1-degree and 3-degree nodes of $R(f)$, and $n_2$ is number 2-degree nodes. As seen previously, number of 1,3-degree nodes and their contours are doubled in $M$. However, number of 2-degree nodes is invariant under doubling, corresponding contours are doubled. \\
Therefore, $R(f_0)$ has $2e$ arcs and $2n_{1,3}+n_2$ nodes. Known fact that number of loops in a graph is $\beta_1 (G)=1-+e-n$, where $e$ is number of edges(arcs) and $n$ is number of nodes. \\
After substitution we get:\\
$1-\beta_1(R(f_0)=2n_{1,2}+n_2-2e=2-2+2n_{1,3}+2n_2+2e-n_2=2-2(1-n_{1,3}-n_2+e)-n_2=2-2\beta_1(R(f))-n_2$. \\
Also we know that number of loops of $N$ is its genus and genus of $M$ (number of loops) is one less that one of $N$. Therefore, we can make a substitution $\beta_1(R(f_0))=g_N-1$. \\
From the above results we get: \\
$\beta_1(R(f))=\frac{g_N-n_2}{2}$.\\
Since number of 2-degree nodes $n_2\geq0 $, we may obtain the following result:\\
\textbf{Lamma ?}\\
The Reeb graph of a Morse function over connected non-orientable 2-manifolds of genus $g$ with boundary is at most $\frac{g}{2}$.\\
